---
title: "CC_borders_cancer_analysis.RMD"
output: html_document
date: '2022-05-07'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(janitor)
library(here)
library(ggplot2)
library(stringr)
```


data loading and preperation
```{r}
incident_hb_5year <- clean_names(read_csv(here("data/5 Year Summary of Incidence by Health Board/e8d33b2b-1fb2-4d59-ad21-20fa2f76d9d5.csv")))
HB_labels <- clean_names(read.csv(here("data/geography_codes_and_labels_hb2014_01042019.csv")))
incident_hb_5year <- incident_hb_5year %>% left_join(HB_labels, "hb")
```
data analysis

```{r}
incident_hb_5year %>% filter(cancer_site == "All cancer types", sex == "All") -> all_region_cancer
  
  ggplot(all_region_cancer) +
  geom_col() + 
  aes(y = crude_rate, x = reorder(hb_name, crude_rate), fill = hb_name == "NHS Borders") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_manual(values = c("blue", "red"), guide = "none") +
  labs(x = "Health Board", 
       y = "Number of Cases (standerdised)"
  )
```


observation: the borders has a higher than avarage rate of cancer per capita, however this is a raw number and not inclusive of age.
```{r}
incident_hb_5year %>% filter(cancer_site == "All cancer types", sex == "All") -> all_region_cancer
  
  ggplot(all_region_cancer) +
  geom_col() + 
  aes(y = standardised_incidence_ratio, x = reorder(hb_name, standardised_incidence_ratio), fill = hb_name == "NHS Borders") +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_manual(values = c("blue", "red"), guide = "none") +
  labs(x = "Health Board", 
       y = "Number of Cases (standerdised)"
  )
```


observation: when standeridiseing for age, the rates are significantly lower relative to the other regions of scotland



```{r}
incident_hb_5year %>%
  rename(incidences_age05to09 = incidences_age5to9) %>%
  rename(incidences_age00to05 = incidences_age_under5) %>%
  filter(hb_name == "NHS Borders", cancer_site == "All cancer types", sex %in% c("Male", "Female")) %>%
  select(cancer_site, sex, starts_with("incidences_age")) %>%
  pivot_longer(-c(cancer_site, sex), names_to = "ages", values_to = "occurances") %>%
  mutate(ages = str_remove_all(ages, "incidences_age")) %>%
  group_by(ages)-> age_data

  ggplot(age_data) +
  aes(x = ages, y = occurances, fill = sex) +
  geom_col(position = "dodge") +
      theme(axis.text.x = element_text(angle = 45)) +
  labs(x = "Age", 
       y = "Number of occurances"
  )

```


observation: cancer in the borders is more prevalent among women however later in life men with the exception of a male spike dureing infancy which may be an anomily


```{r}
incident_hb_5year %>%
 filter(hb_name == "NHS Borders", cancer_site != "All cancer types", sex == "Male") %>% 
  slice_max(incidences_all_ages, n = 10, with_ties = TRUE) %>%
  select(cancer_site, incidences_all_ages) %>% 
  rename(top_10_male = cancer_site, incidences_male = incidences_all_ages) -> top_10

incident_hb_5year %>%
 filter(hb_name == "NHS Borders", cancer_site != "All cancer types", sex == "Female") %>% 
  slice_max(incidences_all_ages, n = 10, with_ties = TRUE) %>%
  select(cancer_site, incidences_all_ages) %>% 
  rename(top_10_female = cancer_site, incidences_female = incidences_all_ages) %>% 
  bind_cols(top_10) -> top_10

top_10
```

```{r}
common <- intersect(top_10$top_10_female, top_10$top_10_male)
common
```


observation summary

the borders has a higher than avarage rate of cancer per capita, however this is a raw number and not inclusive of age.
when standeridiseing for age, the rates are significantly lower relative to the other regions of Scotland.
cancer in the borders is more prevalent among women however later in life men with the exception of a male spike during infancy which may be an anomaly.
as age increases, rates of cancer also increase, until the age of 75 wherein mortality may be significantly reducing the population available to get cancer
men and women in the borders share 7 and have 3 unique cancer diagnosis in their respective top 10 most common cancer diagnosis which could be an indicator for future treatment volumes of specific cancer types.

